Communicative pressures at the semantics-pragmatics interface:\\ Learning biases may prevent the lexicalization of pragmatic inferences

### 2016-10-04 paper/paper.tex
minor changes paper draft

\subsection{Summary}
We argued that expressivity, learnability and simplicity are central to the cultural evolution of language. In the present setup they are modelled as communicative efficiency-relative replication, iterated Bayesian learning, and a prior that biases learners for compressed lexical meanings. Taken together their dynamics are described by the replicator-mutator dynamics \citep{hofbauer+sigmund:2003}:

@@

\subsection{Setup}
 We consider populations with two signaling behaviors, either literal or pragmatic. As mentioned earlier, the former correspond to level $0$ reasoners who only take their lexica into consideration. In the following, pragmatic types correspond to level $1$ reasoners since higher level reasoning is not required to derive scalar implicatures from the lexica fragments we consider here. The prior over states is assumed to be uniform for all behaviors.

The lexica are listed in Table \ref{tab:lexica}. As in our previous examples, they are $(2,2)$-Boolean matrices, i.e., $|M| = |S| = 2$. These are the simplest matrices that allow us to make the presence or lack of an upper-bound and the use of scalar implicatures precise. Lexica $L_1$ to $L_3$ are not optimal for communication because they assign the same meaning to all their messages. They were included to showcase the selection process for a larger hypothesis space. $L_4$ and $L_5$ are our target lexica, codifying upper-bounded semantics for message $m_1$ (the first column in a lexicon) and a lack thereof, respectively. Lastly, $L_6$ is similar to $L_5$ in that two messages are true of the same state but differs from it in assigning upper-bounded semantics to $m_1$. Combining a signaling behavior with one of these $6$ lexica yields a total of $12$ distinct types.\footnote{While there is a total of $16$ possible $(2,2)$-matrices, a number of them are identical both in terms of expressivity and the learning bias against lexical upper-bounds. The competition between such types is determined by their  proportions in the initial population but this fact can be obscured when averaging across simulations. We therefore focus only on the subset that exhibits the properties we set out to explain.}

@@

$L_4$ = $\begin{pmatrix} 1 & 0 \\ 0 & 1 \end{pmatrix}$ &
$L_5$ = $\begin{pmatrix} 1 & 0 \\ 1 & 1 \end{pmatrix}$ &
$L_6$ = $\begin{pmatrix} 1 & 1 \\ 0 & 1 \end{pmatrix}$
\end{tabular}
\caption{Space of possible lexicon fragments considered.}
\label{tab:lexica}
\end{table}

We focus our analysis on the contrast between literal and pragmatic types using lexica $L_4$ and $L_5$. Note in particular  that a type that has conventionalized upper and lower bounds to realize a (quasi-)partition of the relevant semantic space, such as $L_4$, will produce speaker behavior that is {\em almost} indistinguishable from that of a language with only the respective upper bounds, but with Gricean speakers, such as $L_5$. Almost, because there may be slight differences between the probability with which speakers would (erroneously) use a semantically false description and the probability with which speakers would (erroneously) use a pragmatically suboptimal description. This further contrasts with literal $L_5$, a type lacks means to convey an upper-bound with $m_1$. Due to the possible marginal difference in signaling behavior between pragmatic $L_4$ and $L_5$, the selection of one type over the other is expected to mainly depend on the learning bias. Things are less clear for literal $L_5$ contrasted with literal/pragmatic $L_4$. The former has a learning advantage but is expected to fare worse in terms of communicative fitness in virtue of ambiguous $m_1$.

As implicit in the discussion above, one may think of $s_1$ as a ``some but not all''-state and $s_2$ for an ``all''-state. The literal meaning of weak scalar expressions such as {\em some} then corresponds to a message true of both $s_1$ and $s_2$ in these fragments.  Following our assumption for a preference for simple lexical representations, the prior biases learners against lexica in which a message holds true only of the former and not the latter. All other semantics are a priori equally probable. This prior is given by $P(t_i) \propto n - c \cdot r$, where $n$ is the total number of states and $r$ is the number of messages only true of $s_1$ in $t_i$'s lexicon, $c \in [0,1]$. In sum, an increase in $c$ brings about a stronger learning bias against languages that lexicalize upper-bounds, i.e., $L_2, L_4$ and $L_6$.

\subsection{Analysis}
The dynamics are initialized with an arbitrary distribution over types, constituting the population's first generation. If not stated otherwise, the results for a given parameter setting were obtained from $1000$ independent runs. Each run consisted of $20$ generations. This corresponds to a developmental plateau after which no noteworthy change was registered. As specified in \ref{sec:mutation}, the learning transition matrix $Q$ can be obtained by considering all possible state-message sequences of length $k$. Given that this is intractable for large $k$, matrices with $k > 5$ were approximated by using Monte Carlo with $10$ sequences sampled from each type's production probabilities and a type's children being exposed only to this subset. The model's parameters are summarized in Table \ref{tab:summary}.

@@

According to our hypothesis, functional pressure on successful communication combined with learning pressures in the form of a bias against upper-bounds may lead to the selection of $L_5$-like semantics. However, it is instructive to first inspect the effect of these pressures in isolation. For this purpose, we focus on three pragmatic types.\footnote{Pragmatic reasoning allows language users to refine their (possibly erroneous) choices. Therefore, it is advantageous even for those types that codify an upper-bound lexically.} Users of $L_3$, representing a type that is lacking in expressivity but is a prior preferred for its lack of upper-bounds. Users of $L_4$, a type that is functionally advantageous but biased against. And users of $L_5$, combining the virtues of the latter two.

\paragraph{Expressivity only.} Recall that the outcome of the replicator dynamics are influenced by $\lambda$ and $\alpha$ as they have a bearing on a type's fitness. Low $\alpha$ disadvantages types that rely on pragmatic reasoning for more deterministic signaling behavior to the gain of those that codify this information semantically. The rationality parameter $\lambda$ has a similar effect for different reasons. Less utility maximizing behavior decreases the association of an ambiguous message with a single state, even when other states are uniquely associated with a different one. That is, $\lambda$ regulates the strength by which users of $L_5$ associate non-upper-bounded $m_1$ exclusively with the ``some''-state $s_1$ over the ``all''-state $s_2$.

>>>>

\subsection{Summary}
Drawing from past research, we argued that expressivity, learnability and simplicity are central to the cultural evolution of language. In the present setup they are modelled as communicative efficiency-relative replication, iterated Bayesian learning, and a prior that biases learners for compressed lexical meanings. Taken together their dynamics are described by the replicator-mutator dynamics \citep{hofbauer+sigmund:2003}:

@@

\subsection{Analysis}
We analyze the model's predictions in populations of types with two signaling behaviors; literal or pragmatic. The former correspond to level $0$ reasoners who only take their lexica into consideration and the latter to level $1$ reasoners. Higher level reasoning is not required to derive scalar implicatures from the lexica we consider here nor would it provide substantial pragmatic refinement.

The space of possible lexica is given in Table \ref{tab:lexica}. As in the preceding examples, lexica are $(2,2)$-Boolean matrices. These are the simplest matrices that allow us to make the contrast between the presence or lack of an upper-bound and the use of scalar implicatures precise. As implicit in the discussion so far, one may think of $s_1$ as a ``some but not all''-state and of $s_2$ as an ``all''-state. The literal meaning of weak scalar expressions such as English {\em some} then corresponds to a message true of both $s_1$ and $s_2$ in these fragments. Following our assumption for a preference for simple lexical representations, the prior biases learners against lexica in which a message holds true only of state $s_1$. That is, those messages that lexicalize an upper-bound that rules out the ``all''-state $s_2$. All other semantics are assumed to be a priori equally probable. Accordingly, this prior is captured by $P(t_i) \propto n - c \cdot r$, where $n$ is the total number of states and $r$ is the number of messages only true of $s_1$ in $t_i$'s lexicon, $c \in [0,1]$. In short, increments in parameter $c$ bring about a stronger preference against languages that lexicalize upper-bounds, i.e., $L_2, L_4$ and $L_6$.

@@

$L_4$ = $\begin{pmatrix} 1 & 0 \\ 0 & 1 \end{pmatrix}$ &
$L_5$ = $\begin{pmatrix} 1 & 0 \\ 1 & 1 \end{pmatrix}$ &
$L_6$ = $\begin{pmatrix} 1 & 1 \\ 0 & 1 \end{pmatrix}$
\end{tabular}
\caption{Space of possible lexica.}
\label{tab:lexica}
\end{table}

While there is a total of $16$ possible $(2,2)$-lexica, a number of them are identical both in terms of expressivity and the learning bias. The competition between such types is determined by the initial configuration of a population, i.e., the proportion with which each type starts out. However, this fact can be obscured when averaging across simulations. We therefore focus on a smaller representative subset.\footnote{Simulations conducted with the full space of possible lexica confirm that the results reported here do not hinge on their exclusion.} Lexica $L_1$ to $L_3$ are not optimal for communication because they assign the same state to all their messages. This failure to associate a single form to a meaning both semantically and pragmatically inevitably leads to their disadvantage in language use. $L_4$ and $L_5$ are our target lexica, previously labeled as lexica $L_b$ and $L_a$. They codify upper-bounded semantics for message $m_1$ and a lack thereof, respectively. Lastly, $L_6$ is similar to $L_5$ in that two messages are true of the same state but differs from it in assigning upper-bounded semantics to $m_1$.

Combining a signaling behavior with each of these $6$ lexica yields a total of $12$ distinct types. Note in particular  that a type that has conventionalized upper bounds to realize a (quasi-)partition of the relevant semantic space, such as $L_4$, will produce speaker behavior that is {\em almost} indistinguishable from that of a language that lacks upper bounds, but with pragmatic speakers, such as $L_5$. Almost, because there may be slight differences between the probability with which speakers would (erroneously) use a semantically false description and the probability with which speakers would (erroneously) use a pragmatically suboptimal description. Due to this possibly marginal difference between pragmatic $L_4$ and $L_5$, the selection of one type over the other is expected to mainly depend on the learning bias. Things are less clear for literal $L_5$ contrasted with literal/pragmatic $L_4$. The former has a learning advantage but is expected to fare worse in communicative  terms in virtue of ambiguous $m_1$.

The dynamics are initialized with an arbitrary distribution over types, constituting the population's first generation. The results for each parameter setting were obtained from $1000$ independent runs, each consisting of $20$ generations. This corresponds to a developmental plateau after which no noteworthy change was registered. As specified in \S\ref{sec:learnability}, the learning matrix $Q$ can be obtained by considering all possible state-message sequences of length $k$. Given that this is intractable for large $k$, matrices were approximated by sampling $10$ sequences from each type's production probabilities and a type's children being exposed only to this subset. For convenience, the model's parameters are summarized in Table \ref{tab:summary}.

@@

According to our hypothesis, functional pressure on successful communication combined with learning pressures in the form of a bias against upper-bounds may lead to the selection of $L_5$-like semantics. It is instructive to first inspect the effect of these pressures in isolation. For this purpose we focus our attention on three pragmatic types.\footnote{Pragmatic reasoning allows language users to refine their (possibly erroneous) choices. Therefore, it is advantageous even for those types that codify an upper-bound lexically.} Pragmatic $L_3$, a type that is lacking in expressivity but is a priori preferred for its lack of upper-bounds. Pragmatic $L_4$, a type that is functionally advantageous but biased against. And pragmatic $L_5$, combining virtues of the latter two.

\paragraph{Expressivity only.} Recall that the replicator dynamics are sensitive to $\lambda$ and $\alpha$ as both have a bearing on a type's fitness. In particular, low $\alpha$ disadvantages types that rely on pragmatic reasoning to the gain of those that codify more semantically. The rationality parameter $\lambda$ has a similar effect for different reasons: Low rationality leads to a less pronounced preference for the choice(s) expected to succeed best in communication. That is, $\lambda$ regulates the strength by which users of $L_5$ associate non-upper-bounded $m_1$ exclusively with the ``some''-state $s_1$ over the ``all''-state $s_2$. Speakers of $L_4$ need not rely on $\lambda$ for this, as as the association of $m_1$ with $s_1$ is already part of their language's semantics.
